Chatbot based Cognitive Behavioural Therapy

### 2021-05-22 paper.tex
chapter 4.2

\section{Introduction}
During the last two decades the internet has become a major part in the life
of people throughout the society. The ubiquity has led to the desire for 
location-independent usage. As a consequence the amount of
smartphone users is increasing every year.\footnote{https://www.statista.com/statistics/330695/number-of-smartphone-users-worldwide/, last access: 13.05.2021}
With constant access to the biggest collection of knowledge in the world,
more and more individuals tend to use it as a first source of information when
symptoms of illness arise\cite{Wyatt2015}. 
In terms of mental health, the internet provides a numerous amount of \emph{cognitive behavioural therapy} (CBT) offers such as guides
or apps. 
\\\\
In contrast to a complex psychoanalysis with the guidance of a therapist or doctor trying to look into the past,
CBT focuses more on the way an individual deals with its present thoughts and beliefs. The treatment has the aim to
change behavioural patterns in a way that they positively influence emotions and cognition. CBT is based on this connection
between the way how we interpret the daily life, how we behave due to it and how we 
feel in the end which is shown in Figure \ref{fig:cbt_circle}. By changing one of the three aspects,
the others are changed as well automatically. Some popular methods are keeping a \emph{Gratitude Journal} or exercises to calm the body.\cite{Spangler2002}
\\
\emph{Gratitude Journaling} works by writing down moments, situations or facts the writer is grateful for. It can lead to a more positive mindset and higher mindfulness. 
\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{cbt_picture}
  \caption[cbt_circle]{The relationships among cognition, emotion, behavior. Graphic by Raja Selvam, 2018\footnotemark}
  \Description{The relationships among cognition, emotion, behavior.}
  \label{fig:cbt_circle}
\end{figure}
\footnotetext{https://integralsomaticpsychology.com/wp-content/uploads/2018/02/ISP-Embodied-Cognition-Diagram-Raja-Selvam-750.jpg, last access: 30.04.2021}
\\
With the help of apps, users can get hints and instructions to conduct self-help. It is possible
to get first support without the need to consult a therapist or doctor. Often, this a big and 
difficult step for many people due to the fear of getting stigmatized\cite{Rossler2016}.
While CBT applications have the potential to reduce the barrier to seek help, the rate of completion
is relatively low. Donkin et. al \cite{Donkin} found a median minimal completion rate of 56\% during a randomized controlled trial. 
One possible reason for this lack of adherence is the missing human interaction of internet-based CBT\cite{Ly2017}.
The usage of chatbots are considered as a possible solution for this problem. While they are providing CBT on the same level 
as non-chatbot based approaches, they can deliver an artificial human interaction that is supposed to increase the
therapy adherence which can be seen as an important aspect of a successful therapy. 
\\\\
The paper starts with a review of a part of the conducted research and analyses weaknesses and strengths as well as risks and chances. 
Subsequently, it will define some general requirements a chatbot used for CBT should met.
What are aspects that need improvement during further development approaches and what methods and measures might be suitable to achive that? 
Considering the state of the art and the general requirements, the paper tries to answer this question regarding safety aspects, efficiacy and privacy.
It finally discusses the outcomes and points out limits of chatbot based CBT.

\section{Strenghts and weaknesses of existing chatbots}
This far, there are a couple of chatbots available. This paper will mainly focus on the \emph{Woebot} Chatbot and \emph{Wysa} for several reasons.
On the one hand, they have been developed in a research oriented context. 
The examples are available in english language and free to use.\footnote{https://woebothealth.com/about-us/, last access: 01.05.2021}\textsuperscript{,}\footnote{https://www.wysa.io/faq, last access: 01.05.2021}
Additionally, the developers provide numerous information on their website
like experience reports or technical aspects.
On the other hand, a lot of studies have already been conducted mainly focusing on the efficiacy of the chatbot based CBT. 
They revealed some weaknesses of those chatbots, some chances and also risks concerning Efficiacy, Privacy and Safety.

@@

\emph{Woebot} is available as a standalone app in Apple's App Store and in Google Play store where it reached more than 100.000 downloads\footnote{https://play.google.com/store/apps/details?id=com.woebot\&hl=de\&gl=US, last access: 03.05.2021} and as 
a facebook chatbot. It is based on natural language processing and offers free-text inputs as well as predefined answers.
In general, it gives hints how to change the thinking and guides the user through techniques like \emph{Gratitude Journaling}. 

@@

both groups has been found but both showed a reduction of it. Furthermore, the study has revealed that eighty-five percent of persons in the chatbot group used 

@@

is used through them like Facebook's messenger. The access and deletion of it then depends on Facebook's privacy policy.\footnote{https://woebothealth.com/privacy-webview/, last access: 04.05.2021}

@@

Considering emergency situations, \emph{Woebot} is able to recognize user inputs like 'SOS' or 'suicide' as such. It reacts with a list of helplines

@@

in Apple's app store and Google Play store where it has been downloaded more than 1.000.000 times\footnote{https://play.google.com/store/apps/details?id=bot.touchkin, last access: 05.05.2021}. 

@@

with a less frequent usage. In general, both groups showed an improvement. 74\% of the study participants provided an in-app feedback. Most of them preferred the pre-defined answers rather than free-text.

@@

The way \emph{Wysa} stores user data deviates from \emph{Woebot}'s. Though \emph{Wysa} does not require a registration, it links all the collected data to a vendor-specific user id provided by Google Play or App store. This approach leads to a pseudonymised data storage.

@@

It is important to mention, that the chatbot can not be used through third party applications like Facebooks's messenger. Hence a user can use the application more independently of privacy policies of other companies. Altough it is not completely excluded as \emph{Wysa} allows to speak with the chatbot 
via Apple's Siri or Google's voice assistant.\footnote{https://legal.wysa.io/privacy-policy, last access: 08.05.2021}

@@

The chatbot understands words like 'privacy' and can provide links and information about it during a conversation as well.\cite{Kretzschmar2019}

@@

For the second question, the study has revealed that users prefer chatbots with good technical quality like grammatically correct answers rather than those with social aspects like seeing an avatar while chatting.

@@

The conclusion of both results is that users prefer human-like chatbots while having less privacy concerns at the same time during the conversations. Companies could easily exploit this fact. In order not to be accused of it, they should take privacy very serious especially when they are programming chatbots on a professional level.  

@@

to the user in order to raise awareness for problems that can come with intelligent algorithms like false positive classifications.

@@

In order to summarize the privacy requirements of a chatbot for mental health it can be said, that information is the most important one. Firstly, it leads to a conscious usage of the application which is essential to protect personal data. It secondly ensures having control over the data and being able to delete it. Finally, information helps to understand the system and 

@@

>>>>

\section{Introduction}
During the last two decades the internet has become a major part in the life
of people throughout the society. The ubiquity has led to the desire for 
location-independent usage. As a consequence the amount of
smartphone users is increasing every year.\footnote{https://www.statista.com/statistics/330695/number-of-smartphone-users-worldwide/, last access: 13.05.2021}
With constant access to the biggest collection of knowledge in the world,
more and more individuals tend to use it as a first source of information when
symptoms of illness arise\cite{Wyatt2015}. 
In terms of mental health, the internet provides a numerous amount of \emph{cognitive behavioural therapy} (CBT) offers such as guides
or apps. 
\\\\
In contrast to a complex psychoanalysis with the guidance of a therapist or doctor trying to look into the past,
CBT focuses more on the way an individual deals with its present thoughts and beliefs. The treatment has the aim to
change behavioural patterns in a way that they positively influence emotions and cognition. CBT is based on this connection
between the way how we interpret the daily life, how we behave due to it and how we 
feel in the end which is shown in Figure \ref{fig:cbt_circle}. By changing one of the three aspects,
the others are changed as well automatically. Some popular methods are keeping a \emph{Gratitude Journal} or exercises to calm the body.\cite{Spangler2002}
\\
\emph{Gratitude Journaling} works by writing down moments, situations or facts the writer is grateful for. It can lead to a more positive mindset and higher mindfulness. 
\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{cbt_picture}
  \caption[cbt_circle]{The relationships among cognition, emotion, behavior. Graphic by Raja Selvam, 2018\footnotemark}
  \Description{The relationships among cognition, emotion, behavior.}
  \label{fig:cbt_circle}
\end{figure}
\footnotetext{https://integralsomaticpsychology.com/wp-content/uploads/2018/02/ISP-Embodied-Cognition-Diagram-Raja-Selvam-750.jpg, last access: 30.04.2021}
\\
With the help of apps, users can get hints and instructions to conduct self-help. It is possible
to get first support without the need to consult a therapist or doctor. Often, this a big and 
difficult step for many people due to the fear of getting stigmatized\cite{Rossler2016}.
While CBT applications have the potential to reduce the barrier to seek help, the rate of completion
is relatively low. Donkin et. al \cite{Donkin} found a median minimal completion rate of 56\% during a randomized controlled trial. 
One possible reason for this lack of adherence is the missing human interaction of internet-based CBT\cite{Ly2017}.
The usage of chatbots are considered as a possible solution for this problem. While they are providing CBT on the same level 
as non-chatbot based approaches, they can deliver an artificial human interaction that is supposed to increase the
therapy adherence which can be seen as an important aspect of a successful therapy. 
\\\\
The paper starts with a review of the conducted research and analyses weaknesses and strengths as well as risks and chances. 
Subsequently, it will analyse some general requirements a chatbot used for CBT should met.
What are aspects that need improvement during further development approaches and what methods and measures might be suitable to achive that? 
Considering the state of the art and the general requirements, the paper tries to answer this question regarding safety aspects, efficiacy and privacy.
It finally discusses the outcomes and points out limits of chatbot based CBT.

\section{Strenghts and weaknesses of existing chatbots}
This far, there are a couple of chatbots available. This paper will mainly focus on the \emph{Woebot} Chatbot and \emph{Wysa} for several reasons.
On the one hand, they have been developed in a research oriented context. 
The examples are available in english language and free to use.\footnote{https://woebothealth.com/about-us/, last access: 01.05.2021}\textsuperscript{,}\footnote{https://www.wysa.io/faq, last access: 01.05.2021}
\\
Additionally, the developers provide numerous information on their website
like experience reports or technical aspects.
On the other hand, a lot of studies have already been conducted mainly focusing on the efficiacy of the chatbot based CBT. 
They revealed some weaknesses of those chatbots, some chances and also risks concerning Efficiacy, Privacy and Safety.

@@

\emph{Woebot} is available as a standalone app in \emph{Apple}'s App Store and in \emph{Google Play} store where it reached more than 100.000 downloads\footnote{https://play.google.com/store/apps/details?id=com.woebot\&hl=de\&gl=US, last access: 03.05.2021} and as 
a \emph{Facebook} chatbot. It is based on natural language processing and offers free-text inputs as well as predefined answers.
In general, it gives hints how to change the thinking and guides the user through techniques such as \emph{Gratitude Journaling}. 

@@

both groups has been found but both showed a reduction of it. The results have been examined with the help of the PHQ-9 questionnaire for depression symptoms and the GAD-7 questionnaire for anxiety. Furthermore, the study has revealed that 85\% of persons in the chatbot group used 

@@

is used through them like \emph{Facebook's} messenger. The access and deletion of it then depends on \emph{Facebook's} privacy policy.\footnote{https://woebothealth.com/privacy-webview/, last access: 04.05.2021}

@@

Considering emergency situations, \emph{Woebot} is able to recognize user inputs e.g 'SOS' or 'suicide' as such. It reacts with a list of helplines

@@

in \emph{Apple}'s app store and \emph{Google Play} store where it has been downloaded more than 1.000.000 times\footnote{https://play.google.com/store/apps/details?id=bot.touchkin, last access: 05.05.2021}. 

@@

with a less frequent usage. Their participants were human-beings around the globe who have been asked to install the Wysa app. One part of them used the app less often than the other one. O In general, both groups showed an improvement. 74\% of the study participants provided an in-app feedback. Most of them preferred the pre-defined answers rather than free-text.

@@

The way \emph{Wysa} stores user data deviates from \emph{Woebot}'s. Though \emph{Wysa} does not require a registration, it links all the collected data to a vendor-specific user id provided by \emph{Google Play} or App store. This approach leads to a pseudonymised data storage.

@@

It is important to mention, that the chatbot can not be used through third party applications like \emph{Facebook's} messenger. Hence a user can use the application more independently of privacy policies of other companies. Altough it is not completely excluded as \emph{Wysa} allows to speak with the chatbot 
via \emph{Apple}'s Siri or \emph{Google's} voice assistant.\footnote{https://legal.wysa.io/privacy-policy, last access: 08.05.2021}

@@

The chatbot understands words such as 'privacy' and can provide links and information about it during a conversation as well.\cite{Kretzschmar2019}

@@

For the second question, the study has revealed that users prefer chatbots with good technical quality like grammatically correct answers rather than those with social aspects such as seeing an avatar while chatting.

@@

The conclusion of both results is that users prefer human-like chatbots while having less privacy concerns at the same time during the conversations. Companies could easily exploit this fact. In order not to be accused of it, they should focus on privacy especially when they are programming chatbots on a professional level.  

@@

to the user in order to raise awareness for problems that can come with intelligent algorithms such as false positive classifications.

@@

In order to summarize the privacy requirements of a chatbot for mental health it can be said, that information is the most important one. Firstly, it leads to a conscious usage of the application which is essential to protect personal data. Secondly, information is necessary in order to know how data can get deleted or accessed. It finally helps to understand the system and 

@@

After reviewing two chatbots for CBT and analysing requirements for HCI applications in general, some specific development recommendations are now being derivated. 
\subsection{Knowledge and Data Dashboard}
In order to respect the privacy and take user concerns serious, chatbot applications should only exist as a standlone app without the possibility to use it through a third party application such as \emph{Facebook}'s messenger. The dependence of another company's privacy policy excludes control which is a crucial aspect within the 
work of Patrick and Kenny\cite{Patrick2003}. Third party applications can still be used as marketing platforms.
Furthermore, chatbot developers could implement a knowledge and data dashboard. A standalone app could have an extra view which is meant to access and control the data a chatbot has collected.
This approach could have two major advantages. On the one hand, the chatbot user would always be able to get an insight into data that has been gathered. Additionally, those data could be enriched with the chat snippet or user input that has been the source of it. 
This mapping of data and information would probably lead to a better understanding of both concepts in general and how the chatbot infers knowledge in particular. It would contribute to the result of Lee et. al \cite{Lee2011} who noted that users do not distinguish between data and information.
In order to improve transparency basic information about the used algorithm could be provided for users with specific interest or a technical background as well.
On the other hand, it could lead to a more transparent privacy policy and less complex administration processes for the developing companies. Several privacy legislations force companies to give customers access to their stored data. Usually, users have to write an e-mail that needs to be processed by the company. It could save this effort by
implementing the data and knowledge dashboard giving users access to their data constantly.
Moreover, the dashboard could have the option to delete the data. As Saglam et. al \cite{Shamim2021} have shown, chatbot users are worried about how they can delete personal information. Having the possibility to do it directly in the app could lead to a higher trust of the users and again to less complex administration processes for the company
as they would not need to process deletion requests anymore.
The implementation of a knowledge and data dashboard is one possible solution to sharpen the privacy focus. It is also conceivable to integrate the functionalities of the dashboard directly into the chat.
A user input such as 'What do you know about me?' or 'What data have you collected so far?' could lead to an overview of the data in the same manner a dashboard would provide it.
Requests like 'Delete the data!' could then start the wiping process. 
Considering the distinguishing between data and information, questions like 'How did you come to the conclusion that I like rain?' could result in a detailed explaination how the chatbot infered this knowledge.
To summarize the idea of a knowledge and data dashboard it can be said that it has the potential to improve the trust of the users while simplyfing adminstration processes of the company at the same time. 
It would improve the comprehension of the system, giving a human-being control over the data collected by the chatbot and would finally lead to a higher consciousness what kind of personal data is stored.
\subsection{Chatbot Setup Phase}
Chapter 3.2 has presented the PERMA abbrevation by Ludden et. al \cite{Ludden2015} which stands for certain aspects which could improve the efficiacy of a chatbot. A specific development approach which respects PERMA could be the implementation of a quick setup phase prior to the actual chatbot usage.
During it, the chatbot could ask questions that aim to reach a better personalization of the application. Following points could be considered and asked. The Chatbot Setup Phase could be seen as an equivalent to a preliminary talk with a human therapist.
\begin{itemize}
\item\textbf{Age}\\ 
Nowadays, the internet is used by all generations\footnote{http://www.ipsos-mori-generations.com/Internet-and-Technology-Use.html, last access: 22.05.2021}
Therefore, personalization concerning the user age might be useful. Differences between a personalization for younger people and users who are older could be a different set of questions that are asked in order to examine the current mood.
Younger people probably feel more noticed if the chatbot asks about the school routine while questions regarding the workplace should address people of older generations.
\item\textbf{Name}\\
Addressing someone with its name is a quite common personalization approach in current applications and already implemented for Wysa and Woebot. It should be also considered for future developments.
\item\textbf{Design Preferences}\\
While the consideration of the age can directly influence the questions asked during the conversation, aspects such as the design do not influence the dialog but could help to improve the acceptance of the application if the appearance adapts to user preferences. 
\item\textbf{Feedback}\\
A common approach of chatbot based CBT is the recommendation of certain methods. The chatbot could ask after the accomplishment of a method if the individual liked it or not which would be useful knowledge for further method recommendations. It does not fit in the Chatbot Setup Phase but could
be asked during the daily dialogs.
\end{itemize} 
Additionally, too much ambient information should be avoided and the main functionality should be focused. An argument against it could be the funding of the app. The reviewed examples \emph{Woebot} and \emph{Wysa} do not advertise within the application but it is a conceivable approach which leads 
to a trade-off that has to be made. It could be part of future research how advertisements are seen as disturbing ambient information and if they negatively influence the outcomes. 
In order to 
